Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
With feedback Research Seminar

**Include sub-questions?:** How can non-convergence be diagnosed? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?**

@@

Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, use we $AC$ and $\widehat{R}$ to evaluate stationarity and mixing separately, as recommended by e.g., @cowl96, p. 898.

@@

## Simulation Hypothesis (als bullets!!)

@@

The data-generating mechanism is a multivariate normal distribution, representing person data on three predictor variables (from an unspecified social scientific field of study). Let our predictor space be defined by the following a multivariate normal distribution

@@

Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. However, results may not be extrapolated to other missingness mechanisms. **Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.**

>>>>

**Include sub-questions?:** How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?**

@@

Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, use we $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
<!-- @cowl96, p. 898. -->

@@

## Simulation Hypothesis (in bullets!! maybe move to simulation section?)

@@

The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables (from an unspecified social scientific field of study). Let our predictor space be defined by the following a multivariate normal distribution

@@

**Maybe move to discussion:**Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. However, results may not be extrapolated to other missingness mechanisms. **Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.**

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
With feedback Research Seminar

\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
Are common MCMC non-convergence diagnostics appropriate for MICE? And if
so, which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
has not converged? And what are guidelines for practice? Can the
parameter of interest, estimand \(Q\), be correct when the algorithm is
not (yet) converged, and vice versa?}

@@

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, use we \(AC\) and \(\widehat{R}\) to evaluate
stationarity and mixing separately, as recommended by e.g.,
\citet{cowl96}, p.~898.

@@

\hypertarget{simulation-hypothesis-als-bullets}{%
\subsection{Simulation Hypothesis (als
bullets!!)}\label{simulation-hypothesis-als-bullets}}

@@

Proper performance of the convergence diagnostics under at least missing
completely at random (MCAR) missingness mechanism is necessary to
demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as
convergence diagnostics. However, results may not be extrapolated to
other missingness mechanisms. \textbf{Convergence diagnostics should
therefore at least apply to the MCAR situation, before more complex
missingness should be explored.}

>>>>

\textbf{Include sub-questions?:} How can non-convergence be identified
diagnostically? Are common MCMC non-convergence diagnostics appropriate
for MICE? And if so, which threshold should be used to diagnose
non-convergence? How many iterations are sufficient/needed to be able to
diagnose non-convergence? Are the default number of iterations
sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
\textbf{How severe is it when the algorithm has not converged? And what
are guidelines for practice? Can the parameter of interest, estimand
\(Q\), be correct when the algorithm is not (yet) converged, and vice
versa?}

@@

Non-stationarity within chains may be diagnosed with e.g.,
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending on
the precision of estimates. A widely used diagnostic to monitor mixing
between chains is the potential scale reduction factor \(\widehat{R}\)
\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
adaptation, \(\widehat{R}\) might also serve to diagnose
non-stationarity, but this has not yet been thoroughly investigated
\citep{veht19}. Therefore, use we \(\widehat{R}\) and \(AC\) to evaluate
mixing and stationarity separately, as recommended by e.g.,
\citet{cowl96}.

@@

\hypertarget{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}{%
\subsection{Simulation Hypothesis (in bullets!! maybe move to simulation
section?)}\label{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}}

@@

\textbf{Maybe move to discussion:}Proper performance of the convergence
diagnostics under at least missing completely at random (MCAR)
missingness mechanism is necessary to demonstrate the appropriateness of
\(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
may not be extrapolated to other missingness mechanisms.
\textbf{Convergence diagnostics should therefore at least apply to the
MCAR situation, before more complex missingness should be explored.}
